Concurrent Programming in Harmony

### 2021-12-24 book/paper.tex
Pass over Chapter 13

Harmony programs, and indeed the HVM,  manipulate Harmony values.
Harmony values are recursively defined:
they include booleans (<{False}> and <{True}>),
integers (but not floating point numbers),
strings (enclosed by single or double quotes),
sets of Harmony values, and dictionaries
\index{dictionary}%
that map Harmony values to other Harmony values.
Strings that start with a letter or an underscore and only contain
letters, digits, and underscores can be written without quotes by
preceding it with a dot.  So <{.example}> is the same string
as <{"example"}>.

@@

Testing is a way to increase confidence in the correctness
of an implementation.
\autoref{fig:queuedemo} demonstrates how concurrent queues
may be used, but it is not a very thorough test program
for an implementation such as the one in \autoref{fig:queueconc}
and does little to increase our confidence in its correctness.
To wit, if <{get()}> always returned 1, the program
would find no problems.
In this chapter, we will look at approaches to testing
concurrent code.

@@

As with critical sections, when testing a concurrent data structure
we need a specification.
\autoref{fig:queues}(a) shows a sequential specification of a
queue in Harmony.
First we can check if the queue implementation
in \autoref{fig:queueconc} meets the sequential
queue specification in \autoref{fig:queues}(a).
To check if the queue implementation meets the specification,
we need to see if any sequence of queue operations in the
implementation matches a corresponding sequence in the
specification.
We say that the implementation and the specification
have the same \emph{behavior}.
%
\autoref{fig:qtestseq} presents a test program that does
exactly this, for sequences of up to \texttt{NOPS}
queue operations.
It maintains two queues:
\begin{itemize}
\item[] \textit{implq}: the queue of the implementation;
\item[] \textit{specq}: the queue of the specification
\end{itemize}
For each operation, the code first chooses whether to
perform a <{get}> or <{put}> operation.
In the case of a <{put}> operation, the code also
chooses which value to append to the queue.
All operations are performed on both the queue implementation
and the queue specification.
In the case of <{get}>, the results of the operation
on both the implementation and specification are checked against one another.

\emph{Test programs themselves should be tested}.
Just because a test program works with a particular implementation
does not mean the implementation is correct---it may be that the implementation
is incorrect but the test program does not have enough coverage to find any
bugs in the implementation.
So run a test program like this with a variety of queue implementations that
have known bugs in them and make sure that the test program finds them.
\begin{comment}
In the case of testing mutual exclusion with atomic counters, you will want
to see if you can use this method to find the problems in \autoref{fig:naivelock},
\autoref{fig:naiveflags},
and \autoref{fig:naiveturn}.
\end{comment}
Conversely, a test program may be broken in that it finds bugs that do not exist.
\begin{comment}
So, in this case,
you also want to check if an atomic counter solution works with known
correct solutions such as \autoref{fig:peterson} and \autoref{fig:tas}
if possible.
\end{comment}

As with any other test program,
\autoref{fig:qtestseq} may not trigger extant bugs, but
it nonetheless inspires reasonable confidence that the
queue implementation is correct, at least sequentially.
The higher \texttt{NOPS}, the higher the confidence.
It is possible to write similar programs in other languages
such as Python, but the <{choose}> expression in Harmony
makes it relatively easy to explore all corner cases.
%
For example, a common programming mistake is to forget to
update the <{tail}> pointer in <{get()}> in case
the queue becomes empty.  Normally, it is a surprisingly
tricky bug to find.  You can comment out those lines
in \autoref{fig:queueconc} and run the test program---it should
easily find the bug and explain exactly how
the bug manifests itself, adding confidence that the test program
is reasonably thorough.

The test program also finds some common mistakes in using
locks, such as forgetting to release a lock when the queue
is empty, but it is not designed to find concurrency bugs
in general.  If you remove all \texttt{acquire()} and
\texttt{release()} calls from \autoref{fig:queueconc},
the test program will not (and should not)
find any errors.

@@

We would like a way that systematically compares behaviors of the
concurrent queue implementation with behaviors of the concurrent
queue specification.  But we cannot do this by composing the specification
and the implementation and simply run the same test operations on both as we
did before---concurrency make the operations non-determistic and thus
the specification and implementation of a single execution might produce
different results, even if both are correct.
Instead, we will create a test program that tries various concurrent
combinations of queue operations, and run it twice: once against the
specification of the concurrent queue and once against the
implementation.  We will then check if the behaviors obtained from
running the implementation is a subset of the behaviors obtained
from the specification.

We will start with a test program that tries concurrent combinations
of various queue operations.
\autoref{fig:qtestpar} shows the test program.  It starts
\texttt{NOPS} threads and each choose to do either a <{put}>
or a <{get}> operation.  In case of a <{put}> operation,
the thread enqueues its own name.  In order to output the behaviors,
each thread prints what operation it is about to perform,
and afterwards it prints that the operation has
completed (including the return value if any).
The figure shows the deterministic finite automaton when the test
program is run against the specification in the case
$\texttt{NOPS} = 2$---for $\texttt{NOPS} = 4$ it would be much
too large to print here.  Since there are no cycles in the DFA, you
can follow some paths through this DFA and see that they are valid
interleavings of the threads.

If you run the same test program against the implementation
of \autoref{fig:queueconc}, you will get the same output.  You can
try this for various $\texttt{NOPS}$, although it gets increasingly
harder to check by hand that the generated DFAs are the same as
\texttt{NOPS} increases.
Now run the test program against \autoref{fig:queueseq}, which is
clearly an incorrect implementation of the concurrent queue specification
because it contains no synchronization code.
However, Harmony does not immediately detect
any problems.  In fact, for $\mathtt{NOPS} = 2$ it even generates
the same set of behaviors.
This is because the test program only outputs the behaviors---it does
not check if they are correct.

Harmony does have a way to check the behavior of one program against
the behavior of another.  In particular, we want to check if the
behaviors of the implementations we have match the behavior of the
specification.  The following shows, for example, how to check the
\texttt{queueconc.hny} implementation on the command line:

@@

\section*{Exercises}
\begin{problems}
\item Write a sequential specification for \autoref{fig:linkedlist}.
(Hint: it implements a set of integers.)
Write a Harmony program that checks if \autoref{fig:linkedlist} meets
the sequential specification.
\item Also create a concurrent specification and a concurrent test program.
Check if the behaviors of \autoref{fig:linkedlist} match those of the
specification.
% \item Extend \autoref{fig:linkedlist} with linearization points and
% check if the implementation is linearizable.
% \item Check if the extended queue implementation of \autoref{ex:qremove}
% is linearizable.
\item Rewrite \autoref{fig:qtestseq} so it only imports \texttt{queue}
and runs \texttt{NOPS} nondeterministically chosen operations against
it (similar in style to \autoref{fig:qtestpar} but without threads).
Then use behaviors to check that \autoref{fig:queueconc} and
\autoref{fig:queueseq} are correct sequential implementations of
the queue.  Check your test program by also trying it on one or two
buggy queue implementations.
\end{problems}

@@

So you wrote a Harmony program and Harmony reports a bug.
Often you may just be able to figure it out by staring at the code and
going through some easy scenarios, but what if you don't?
The output of Harmony can be helpful in that case.

@@

Before we go look at the details of what went wrong, we want to make
sure we generate the simplest scenario.  So first we want to explore
what the smallest \texttt{NOPS} (number of operations or number of
threads) that causes the bug to occur.  With some experimentation,
we find that \texttt{NOPS} = 2 does not find a problem, but
\texttt{NOPS} = 3 does
(\texttt{harmony -m queue=queuebroken -c NOPS=3 code/qtestpar.hny})).
\autoref{fig:queuebroken1} shows the Harmony output.

@@

So how did we get there?  In the top we can see that the order of
events was the following:
\begin{enumerate}
\item[0:] initialization completed, with <{q}> being
\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \};
\item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
<{put}>(1) (see the output column for that clue: the thread printed that).
We can see that <{q}>.<{next}> now points to \texttt{alloc}.\textit{pool}[0],
which the thread must have allocated.  The contents is 
\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: 1 \}, as expected;
\item[2:] thread T2 (\texttt{thread}(1)) started running, calling
<{put}>(?<{q}>, 2).  We can see it got as far as putting 2 on the queue,
but it is not yet done.  It is currently trying to acquire
\texttt{alloc}.\textit{pool}[0].\texttt{lock};
\item[3:] thread T3 (\texttt{thread}(1)) started running, calling
<{get}>($?q$).  We can also see that it freed \textit{pool}[0],
and is now releasing <{q}>.\texttt{lock};
\item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
no longer exists (because T3 just freed it).
\end{enumerate}

@@

Each replica maintains its history and a set of its predecessors.
It then loops until either it fails or it has applied all
\texttt{NOPS} operations to its history.
$\mathbf{choose}(\{ \mathit{immortal}, <{True}> \})$ can only
evaluate to <{False}> in case $\textit{immortal}$ is false; otherwise
it will always evaluate to <{True}>.
What it does within the loop depends on whether it is the head
(has no alive predecessors) or not.
Because failure detection is accurate, at most one replica can think
it is the head at any time (and, if so, it is in fact the head).
Moreover, when it has
detected all its predecessors having failed, eventually some replica
thinks it is the head.

@@

If a replica thinks it is the head, it adds an operation to its
history.
In practice this would be some operation that the head would have received
from a client.  Clients send their operations to the head, which adds them
to its history in the order received.
We are not modeling clients of the service.
Instead, in the model, the head adds its identifier \textit{self} to the
history until the
history has \texttt{NOPS} operations on it or until the head crashes,
whichever comes first.  The head also broadcasts each update, which
are intended for its successor.  Because of failures, it may be that
at different times a replica has different successors, hence the
broadcast.

@@

When the replica successfully adds all \texttt{NOPS} updates to its
history without crashing, the code makes sure that any other such replica
ends in the same state.  Variable \textit{final} is used to check the
different histories against one another.

@@

In the first phase, a client asks each server for its current
timestamp and value.  After receiving $\mathtt{N} - \mathtt{F}$ responses,
the client determines the response with the highest timestamp.
In case of a write operation, the client then computes a new unique
timestamp that is strictly higher than the highest it has seen.
To make this work, timestamps are actually lexicographically ordered
tuples consisting of an integer and the unique identifier of the
client that writes the value.  So if $(t, c)$ is the highest timestamp
observed by client $c'$, and $c'$  needs to create a new timestamp,
it can select $(t + 1, c')$.
After all $(t + 1, c') > (t, u)$ and no other client
will create the same timestamp.

@@

While one can run this code within little time for $\mathtt{F} = 2$, for
$\mathtt{F} = 3$ the state space to explore is already quite large.
One way to reduce the state space to explore is the following realization:
each processor only considers messages for the round that it is in.
If a message is for an old round, the processor will ignore it;
if a message is for a future round, the processor will buffer it.
So one can simplify the model and have each processor wait
for \emph{all} <{N}> messages in a round
instead of $\mathtt{N} - \mathtt{F}$.
It would still have to choose to consider just $\mathtt{N} - \mathtt{F}$
out of those <{N}> messages, but executions in which some processors
are left behind in all rounds are no longer considered.
It still includes executions where some subset of $\mathtt{N} - \mathtt{F}$
processors only choose each other messages and essentially ignore the
messages of the remaining \texttt{F} processors, so the resulting model
is just as good.

@@

\begin{enumerate}
\item Alice creates a new nonce $N_A$ and sends $\langle 1, A, N_A \rangle_\mathtt{Corey}$
to Corey;
\item Upon receipt, Corey sends $\langle 1, A, N_A \rangle_\mathtt{Bob}$ to Bob;
\item Upon receipt, Bob, believing it is engaging in the protocol with Alice,
creates a new nonce $N_B$ and sends
$\langle 2, N_A, N_B \rangle_\mathtt{Alice}$ to Alice;
\item Alice thinks the message came from Corey (because it contains $N_A$, which
Alice created for Corey and sent to Corey) and
sends $\langle 3, N_B \rangle_\mathtt{Corey}$ to Corey.
\item Corey learns $N_B$ and sends $\langle 3, N_B \rangle_\mathtt{Bob}$ to Bob.
\item Bob receiving $\langle 3, N_B \rangle_\mathtt{Bob}$ is identical to the
situation in which Alice tried to set up a connection to Bob,
so Bob now thinks it is talking to Alice, even though Alice never tried to
communicate with Bob.
\end{enumerate}
The security property is violated.  In particular, Bob, duped by Corey, finished
the protocol with Alice and received $A_N$, and even though Bob and Alice
are both honest, Corey has a copy of $A_N$.  So Corey is now able to impersonate
Alice to Bob (but not vice versa because Alice did not try to authenticate Bob).

@@

Charm has some tricks to significantly reduce the state space to explore.
\begin{itemize}
\item A thread can have local state (program counter, stack,
method variables, and thread-local state variables).
That state is called the \emph{context} of the thread.
The context of a thread cannot be accessed by other
threads, nor by \textbf{invariant} statements.
So the model checker only interleaves threads at \textbf{Load},
\textbf{Store}, and \textbf{Del} instructions where a thread
interacts with global
variables.
\item Threads are anonymous, and therefore two or more threads can
have the same context.  The state of the model checker therefore maintains
a \emph{bag} (multiset) of contexts rather are than a \emph{set} of contexts.
Thus even if there are hundreds of threads, there may be only tens of
possible context states.
\end{itemize}

@@

In the output, each thread has a unique identifier:
\texttt{T0} is the initialization thread; \texttt{T}$n$ is the $n^{th}$
spawned thread that executes.  This seems to contradict the fact that
Harmony threads are anonymous.  The output analyzer assigns these
identifiers \emph{a posteriori} to the threads in the state graph
by keeping track, along the reported execution path, what state each
thread is in.  So by examining the initial context of the thread that
is running from some particular state, it can determine if that context
corresponds to the current context of some thread that ran previously
or if the context belongs to a new thread that has not run before.

@@

\texttt{Harmony IDE} is an Integrated Development Environment for Harmony,
also developed by Kevin Sun and Anthony Yang,
that runs in the cloud.  An important advantage of using \texttt{Harmony IDE}
is that you do not have to install anything.  You can even run in on your
smartphone.  However, since Harmony programs can use significant compute
resources, \texttt{Harmony IDE} must put a limit on the amount of compute
resources you can use.  So while \texttt{Harmony IDE} is currently a great
way to try out Harmony, if you become a serious user of Harmony you
will probably want to install either the command-line version or
\texttt{HarmonyLang}.

>>>>

Harmony programs, and indeed the HVM,  manipulate Harmony values.
Harmony values are recursively defined:
they include booleans (<{False}> and <{True}>),
integers (but not floating point numbers),
strings (enclosed by single or double quotes),
sets of Harmony values, and dictionaries
\index{dictionary}%
that map Harmony values to other Harmony values.
Strings that start with a letter or an underscore and only contain
letters, digits, and underscores can be written without quotes by
preceding it with a dot.  So, <{.example}> is the same string
as <{"example"}>.

@@

Testing is a way to increase confidence in the correctness
of an implementation.
\autoref{fig:queuedemo} demonstrates how concurrent queues
may be used, but it is not a very thorough test program
for an implementation such as the one in \autoref{fig:queueconc}
and does little to increase our confidence in its correctness.
To wit, if <{get()}> always returned 1, the program
would find no problems.
Similarly, \autoref{fig:intsettest} is not a good test program
for something as complicated as \autoref{fig:linkedlist}.
In this chapter, we will look at approaches to testing
concurrent code.

@@

As with critical sections---when testing a concurrent data structure---we
need a specification.
For example, \autoref{fig:queues}(a) shows a sequential specification of a
queue in Harmony.
First, we can check if the queue implementation
in \autoref{fig:queueconc} meets the sequential
queue specification in \autoref{fig:queues}(a).
To check if the queue implementation meets the specification,
we need to see if any sequence of queue operations in the
implementation matches a corresponding sequence in the
specification.
We say that the implementation and the specification
have the same \emph{behavior}.
%
\autoref{fig:qtestseq} presents a test program that does
exactly this, for sequences of up to <{NOPS}>
queue operations.
It maintains two queues:
\begin{itemize}
\item[] \textit{specq}: the queue of the specification;
\item[] \textit{implq}: the queue of the implementation.
\end{itemize}
For each operation, the code first chooses whether to
perform a <{get}> or <{put}> operation.
In the case of a <{put}> operation, the code also
chooses which value to append to the queue.
All operations are performed on both the queue implementation
and the queue specification.
In the case of <{get}>, the results of the operation
on both the implementation and specification are checked against one another.

\emph{Test programs themselves should be tested}.
Just because a test program works with a particular implementation
does not mean the implementation is correct---it may be that the implementation
is incorrect but the test program does not have enough coverage to find any
bugs in the implementation.
So, run a test program like this with a variety of queue implementations that
have known bugs in them and make sure that the test program finds them.
\begin{comment}
In the case of testing mutual exclusion with atomic counters, you will want
to see if you can use this method to find the problems in \autoref{fig:naivelock},
\autoref{fig:naiveflags},
and \autoref{fig:naiveturn}.
\end{comment}
Conversely, a test program may be broken in that it finds bugs that do not exist.
In my experience, it is often harder to implement the test program than
the algorithm that the test program tests.

\begin{comment}
So, in this case,
you also want to check if an atomic counter solution works with known
correct solutions such as \autoref{fig:peterson} and \autoref{fig:tas}
if possible.
\end{comment}

As with any other test program,
\autoref{fig:qtestseq} may not trigger extant bugs, but
it nonetheless inspires reasonable confidence that the
queue implementation is correct, at least sequentially.
The higher <{NOPS}>, the higher the confidence.
It is possible to write similar programs in other languages
such as Python, but the <{choose}> expression in Harmony
makes it relatively easy to explore all corner cases.
%
For example, a common programming mistake is to forget to
update the <{tail}> pointer in <{get()}> in case
the queue becomes empty.  Normally, it is a surprisingly
tricky bug to find.  You can comment out those lines
in \autoref{fig:queueconc} and run the test program---it should
easily find the bug and explain exactly how
the bug manifests itself, adding confidence that the test program
is reasonably thorough.

The test program also finds some common mistakes in using
locks, such as forgetting to release a lock when the queue
is empty, but it is not designed to find concurrency bugs
in general.  If you remove all \texttt{acquire()} and
\texttt{release()} calls from \autoref{fig:queueconc},
the test program will not (and should not)
find any errors, but it would be an incorrect implementation
of a concurrent queue.

@@

We would like a way that---similar to the sequential test---systematically
compares behaviors of the
concurrent queue implementation with behaviors of the concurrent
queue specification.  But we cannot do this by composing the specification
and the implementation and simply run the same test operations on both as we
did before---concurrency make the operations non-determistic and thus
the specification and implementation of a single execution might produce
different results, even if both are correct.
Instead, we will create a test program that tries various concurrent
combinations of queue operations, and run it twice: once against the
specification of the concurrent queue and once against the
implementation.  We will then check if the behaviors obtained from
running the implementation are also behaviors obtained
from the specification.

We will start with a test program that tries concurrent combinations
of various queue operations.
\autoref{fig:qtestpar} shows the test program.  It starts
<{NOPS}> threads, and each choose to do either a <{put}>
or a <{get}> operation.  In case of a <{put}> operation,
the thread enqueues its own name (which is provided as an argument
to the thread).  In order to capture the behaviors,
each thread prints what operation it is about to perform,
and afterwards it prints that the operation has
completed (including the return value if any).
This is probably much like you would do if you were trying to find a
bug in a program.

\autoref{fig:qtestpar} also shows the deterministic finite automaton
that describes the possible outputs when the test
program is run against the specification in the case
$<{NOPS}> = 2$---for $<{NOPS}> = 4$ it would be much
too large to print here.  Since there are no cycles in the DFA, you
can follow some paths through this DFA and see that they are valid
interleavings of the threads.
You can obtain this output yourself by running

\begin{code}
\begin{verbatim}
$ harmony -c NOPS=2 -o spec.png code/qtestpar.hny
\end{verbatim}
\end{code}

If you run the same test program against the implementation
of \autoref{fig:queueconc}, you will get the same output:

\begin{code}
\begin{verbatim}
$ harmony -c NOPS=2 -o impl.png -m queue=queueconc code/qtestpar.hny
\end{verbatim}
\end{code}

You can
try this for various $<{NOPS}>$, although it gets increasingly
harder to check by hand that the generated DFAs are the same as
<{NOPS}> increases.
Now run the test program against \autoref{fig:queueseq}, which is
clearly an incorrect implementation of the concurrent queue specification
because it contains no synchronization code.
However, Harmony does not immediately detect
any problems.  In fact, for $\mathtt{NOPS} = 2$ it even generates
the same set of behaviors.
This is because the test program only outputs the behaviors---it does
not check if they are correct.

Harmony does have a way to check the behaviors of one program against
the behaviors of another.  In particular, we want to check if the
behaviors of the implementations we have match behaviors of the
specification.  The following shows, for example, how to check the
\texttt{queueconc.hny} implementation on the command line:

@@

\section*{Exercises}
\begin{problems}
\item Write a Harmony program that checks if
\autoref{fig:linkedlist} satisfies the specification
of \autoref{fig:setobj} \emph{sequentially}.
\item Write a Harmony program that checks if
\autoref{fig:linkedlist} satisfies the specification
of \autoref{fig:setobj} \emph{concurrently}.
% \item Extend \autoref{fig:linkedlist} with linearization points and
% check if the implementation is linearizable.
% \item Check if the extended queue implementation of \autoref{ex:qremove}
% is linearizable.
\item Rewrite \autoref{fig:qtestseq} so it only imports \texttt{queue}
and runs <{NOPS}> nondeterministically chosen operations against
it (similar in style to \autoref{fig:qtestpar} but without threads).
Then use behaviors to check that \autoref{fig:queueconc} and
\autoref{fig:queueseq} are correct sequential implementations of
the queue.  Check your test program by also trying it on one or two
buggy queue implementations.
\end{problems}

@@

So, you wrote a Harmony program and Harmony reports a bug.
Often you may just be able to figure it out by staring at the code and
going through some easy scenarios, but what if you don't?
The output of Harmony can be helpful in that case.

@@

Before we go look at the details of what went wrong, we want to make
sure we generate the simplest scenario.  So, first we want to explore
what the smallest <{NOPS}> (number of operations or number of
threads) that causes the bug to occur.  With some experimentation,
we find that <{NOPS}> = 2 does not find a problem, but
<{NOPS}> = 3 does
(\texttt{harmony -m queue=queuebroken -c NOPS=3 code/qtestpar.hny})).
\autoref{fig:queuebroken1} shows the Harmony output.

@@

So, how did we get there?  In the top we can see that the order of
events was the following:
\begin{enumerate}
\item[0:] initialization completed, with <{q}> being
\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: <{None}> \};
\item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
<{put}>(1) (see the output column for that clue: the thread printed that).
We can see that <{q}>.<{next}> now points to \texttt{alloc}.\textit{pool}[0],
which the thread must have allocated.  The contents is 
\{ \texttt{.lock}: <{False}>, \texttt{.next}: <{None}>, \texttt{.value}: 1 \}, as expected;
\item[2:] thread T2 (\texttt{thread}(1)) started running, calling
<{put}>(?<{q}>, 2).  We can see it got as far as putting 2 on the queue,
but it is not yet done.  It is currently trying to acquire
\texttt{alloc}.\textit{pool}[0].\texttt{lock};
\item[3:] thread T3 (\texttt{thread}(1)) started running, calling
<{get}>($?q$).  We can also see that it freed \textit{pool}[0],
and is now releasing <{q}>.\texttt{lock};
\item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
no longer exists (because T3 just freed it).
\end{enumerate}

@@

Each replica maintains its history and a set of its predecessors.
It then loops until either it fails or it has applied all
<{NOPS}> operations to its history.
$\mathbf{choose}(\{ \mathit{immortal}, <{True}> \})$ can only
evaluate to <{False}> in case $\textit{immortal}$ is false; otherwise
it will always evaluate to <{True}>.
What it does within the loop depends on whether it is the head
(has no alive predecessors) or not.
Because failure detection is accurate, at most one replica can think
it is the head at any time (and, if so, it is in fact the head).
Moreover, when it has
detected all its predecessors having failed, eventually some replica
thinks it is the head.

@@

If a replica thinks it is the head, it adds an operation to its
history.
In practice this would be some operation that the head would have received
from a client.  Clients send their operations to the head, which adds them
to its history in the order received.
We are not modeling clients of the service.
Instead, in the model, the head adds its identifier \textit{self} to the
history until the
history has <{NOPS}> operations on it or until the head crashes,
whichever comes first.  The head also broadcasts each update, which
are intended for its successor.  Because of failures, it may be that
at different times a replica has different successors, hence the
broadcast.

@@

When the replica successfully adds all <{NOPS}> updates to its
history without crashing, the code makes sure that any other such replica
ends in the same state.  Variable \textit{final} is used to check the
different histories against one another.

@@

In the first phase, a client asks each server for its current
timestamp and value.  After receiving $\mathtt{N} - \mathtt{F}$ responses,
the client determines the response with the highest timestamp.
In case of a write operation, the client then computes a new unique
timestamp that is strictly higher than the highest it has seen.
To make this work, timestamps are actually lexicographically ordered
tuples consisting of an integer and the unique identifier of the
client that writes the value.  So, if $(t, c)$ is the highest timestamp
observed by client $c'$, and $c'$  needs to create a new timestamp,
it can select $(t + 1, c')$.
After all $(t + 1, c') > (t, u)$ and no other client
will create the same timestamp.

@@

While one can run this code within little time for $\mathtt{F} = 2$, for
$\mathtt{F} = 3$ the state space to explore is already quite large.
One way to reduce the state space to explore is the following realization:
each processor only considers messages for the round that it is in.
If a message is for an old round, the processor will ignore it;
if a message is for a future round, the processor will buffer it.
So, one can simplify the model and have each processor wait
for \emph{all} <{N}> messages in a round
instead of $\mathtt{N} - \mathtt{F}$.
It would still have to choose to consider just $\mathtt{N} - \mathtt{F}$
out of those <{N}> messages, but executions in which some processors
are left behind in all rounds are no longer considered.
It still includes executions where some subset of $\mathtt{N} - \mathtt{F}$
processors only choose each other messages and essentially ignore the
messages of the remaining \texttt{F} processors, so the resulting model
is just as good.

@@

\begin{enumerate}
\item Alice creates a new nonce $N_A$ and sends $\langle 1, A, N_A \rangle_\mathtt{Corey}$
to Corey;
\item Upon receipt, Corey sends $\langle 1, A, N_A \rangle_\mathtt{Bob}$ to Bob;
\item Upon receipt, Bob, believing it is engaging in the protocol with Alice,
creates a new nonce $N_B$ and sends
$\langle 2, N_A, N_B \rangle_\mathtt{Alice}$ to Alice;
\item Alice thinks the message came from Corey (because it contains $N_A$, which
Alice created for Corey and sent to Corey) and
sends $\langle 3, N_B \rangle_\mathtt{Corey}$ to Corey.
\item Corey learns $N_B$ and sends $\langle 3, N_B \rangle_\mathtt{Bob}$ to Bob.
\item Bob receiving $\langle 3, N_B \rangle_\mathtt{Bob}$ is identical to the
situation in which Alice tried to set up a connection to Bob,
so Bob now thinks it is talking to Alice, even though Alice never tried to
communicate with Bob.
\end{enumerate}
The security property is violated.  In particular, Bob, duped by Corey, finished
the protocol with Alice and received $A_N$, and even though Bob and Alice
are both honest, Corey has a copy of $A_N$.  So, Corey is now able to impersonate
Alice to Bob (but not vice versa because Alice did not try to authenticate Bob).

@@

Charm has some tricks to significantly reduce the state space to explore.
\begin{itemize}
\item A thread can have local state (program counter, stack,
method variables, and thread-local state variables).
That state is called the \emph{context} of the thread.
The context of a thread cannot be accessed by other
threads, nor by \textbf{invariant} statements.
So, the model checker only interleaves threads at \textbf{Load},
\textbf{Store}, and \textbf{Del} instructions where a thread
interacts with global
variables.
\item Threads are anonymous, and therefore two or more threads can
have the same context.  The state of the model checker therefore maintains
a \emph{bag} (multiset) of contexts rather are than a \emph{set} of contexts.
Thus even if there are hundreds of threads, there may be only tens of
possible context states.
\end{itemize}

@@

In the output, each thread has a unique identifier:
\texttt{T0} is the initialization thread; \texttt{T}$n$ is the $n^{th}$
spawned thread that executes.  This seems to contradict the fact that
Harmony threads are anonymous.  The output analyzer assigns these
identifiers \emph{a posteriori} to the threads in the state graph
by keeping track, along the reported execution path, what state each
thread is in.  So, by examining the initial context of the thread that
is running from some particular state, it can determine if that context
corresponds to the current context of some thread that ran previously
or if the context belongs to a new thread that has not run before.

@@

\texttt{Harmony IDE} is an Integrated Development Environment for Harmony,
also developed by Kevin Sun and Anthony Yang,
that runs in the cloud.  An important advantage of using \texttt{Harmony IDE}
is that you do not have to install anything.  You can even run in on your
smartphone.  However, since Harmony programs can use significant compute
resources, \texttt{Harmony IDE} must put a limit on the amount of compute
resources you can use.  So, while \texttt{Harmony IDE} is currently a great
way to try out Harmony, if you become a serious user of Harmony you
will probably want to install either the command-line version or
\texttt{HarmonyLang}.
